TBD

### 2020-08-21 Articles/2019-Entropy/Source/entropy-article.tex
Up to entropy maximization

The overall goal is to describe a generic system as it evolves in time by defining a minimal set of concepts that such a setup must define. We want to find notions that can be applied, for example, to thermodyamic and dynamical systems alike. Then we want to derive general results based on such definitions. Here we provide an overview of the main features, leaving the formal definitions in the (yet to be written) appendix.

@@

Our starting point is the \textbf{process domain} $\tdomain[P]$, which is the set that contains all possible descriptions of the system at all levels of detail at all possible times. Depending on the process, it will contain statements like \statement{the average volume between 1 and 2 seconds is between 3 and 4 liters} or \statement{the trajectory of the particle is $y=10 \, m - t^2 \cdot 9.80665 \, m/s^2 $}. It does not matter at this point what the statements themselves are, except they must be, at least in line of principle, defined from experimentally well-defined starting points. Mathematically, $\tdomain[P]$ is a $\sigma$-complete Boolean algebra\footnote{A set of statements that is closed under negation (NOT), countable disjunction (OR) and countable conjunction (AND).} and it is the closure of the set of experimentally verifiable statements $\edomain[P] \subseteq \tdomain[P]$.\footnote{The set $\edomain[P]$ will form a Heyting algebra which, to be experimentally reachable, must allow a countable basis.}

@@

@@

\begin{table}[h!]
	\centering
\begin{tabular}[h]{|c|c|p{6cm}|}
	\hline 
	Symbol & Name & Meaning \\ 
	\hline 
	$\tdomain[P]$ & Process domain & The set of all possible descriptions at all times and at all levels of granularity \\ 
	\hline 
	$E \subset \tdomain[P]$ & Possible evolutions & The set of descriptions that give a full account of the process; they correspond to the narrowest statements within $\tdomain[P]$ \\ 
	\hline 
	$A(\stmt)$ & Compatible evolutions & The set of evolutions that are compatible with the statement $\stmt$ \\ 
	\hline 
	$\narrower$ & narrower & A statement is narrower than another $\stmt_1 \narrower \stmt_2$ if the second one is true whenever the first one is \\ 
	\hline 
	$\comp$ & compatible & A statement is compatible with another $\stmt_1 \comp \stmt_2$ if they can be true at the same time \\ 
	\hline 
\end{tabular} 
	\caption{Process descriptions and their logical relationships}
	\label{table:logic}
\end{table}

@@

As the process extends over time, we want to organize the statements in our process domain into partial domains, one for each moment in time. Given a \textbf{time parameter} $t \in T \subseteq \mathbb{R}$, we can imagine to carve a subdomain $\tdomain_t$ that would correspond to all the statements about the system one can make at the given time. Note that $\tdomain_t$ is a $\sigma$-complete Boolean sub-algebra of $\tdomain[P]$ since any logical operation on statements at fixed $t$ is also another statement at fixed $t$. The set $X_t \subset \tdomain_t$ of the narrowest statements will give the most precise description at that time, we call these the  possible \textbf{snapshots} of our system at that time.

@@

We should also note that the set of possible snapshots $X_t$ may not be the same at all times and may not correspond to the state of the system. First of all $X_t$ will only include those configurations that are compatible with least by one possible evolution (i.e. $A(x) \neq \emptyset$ for all $x \in X_t$), which may change in time. For example, the set of possible configurations for a system under dissaptive processes will shrink as they converge to equilibrium. In quantum mechanics, the possible configurations after a measurement are restricted to the eigenestates, and a different choice of measurement will lead to a different process with different possible evolutions. Secondly, the best possible description may be broader than the full state of the system (e.g. when external interference forces us to give only a statistical account within our $\Delta t$) or narrower (e.g. when there are known correlations to the other degrees of freedom of the environment or of the system itself). The idea of a state space of the system, independent of the process, is something we need to construct. Finding the requirements needed for that construction, the hidden assumptions, is precisely the issue we need to address.

We say a process is \textbf{deterministic} over the time domains $\{X_t\}_{t \in T}$ if given the snapshot at one time we can always predict the snapshot at all future times. Formally, let $t_1, t_2 \in T$ such that $t_1 \leq t_2$, for any $x_1 \in X_{t_1}$ we can find an $x_2 \in X_{t_2}$ such that $x_1 \narrower x_2$.

@@

Conversely, we say a process is \textbf{reversible} over the time domains $\{X_t\}_{t \in T}$ if given the snapshot at one time we can always reconstruct the snapshot at all past times. Formally, let $t_1, t_2 \in T$ such that $t_1 \leq t_2$, for any $x_2 \in X_{t_2}$ we can find a $x_1 \in X_{t_1}$ such that $x_2 \narrower x_1$.

@@

If we limit ourselves to deterministic processes, with these few definitions, we have already found something that ``grows bigger'' under irreversible evolution while it remains the ``the same size'' under reversible evolution.

@@

\begin{table}[h!]
	\centering
	\begin{tabular}[h]{|c|c|p{6cm}|}
		\hline 
		Symbol & Name & Meaning \\ 
		\hline 
		$t \in T$ & Time parameter & Time is treated as a real valued parameter within the range $T \subseteq \mathbb{R}$ \\ 
		\hline 
		$\tdomain_t \subset \tdomain[P]$ & Time domain & The set of descriptions at a given time \\ 
		\hline 
		$X_t \subset \tdomain_t$ & Possible states & The set of descriptions that give a full account at time $t$; they correspond to the narrowest statements within $\tdomain_t$ \\ 
		\hline 
		$x(t) \narrower x(t + \Delta t)$ & Determinism & Given the state at one time we can predict the state at future times \\ 
		\hline 
		$x(t) \narrower x(t - \Delta t)$ & Reversibility & Given the state at one time we can reconstruct the state at past times \\ 
		\hline 
	\end{tabular} 
	\caption{Time domain, states and determinism}
	\label{table:states}
\end{table}

@@

Quantifying evolutions allows us to define probability as well. If we consider all the evolutions compatible with \statement{the coin is fair}, we must have that half of them are compatible with \statement{the coin is head} while the other are compatible with \statement{the coin is tails}.  Therefore we have $P(\stmt_2 | \stmt_1) = \mu_{\stmt_1}(\stmt_1 \AND \stmt_2)$ where $\stmt_1 \AND \stmt_2$ is the logical conjunction (AND) between the two statements. In other words, the probability quantifies the ratio of evolutions compatible with $\stmt_1$ that are also compatible with $\stmt_2$.

@@

\begin{table}[h!]
	\centering
	\begin{tabular}[h]{|c|c|p{6cm}|}
		\hline 
		Symbol & Name & Meaning \\ 
		\hline 
		$\finer$ & finer & A statement is finer than another $\stmt_1 \finer \stmt_2$ if it provides a more refined description than the second \\ 
		\hline 
		$\eqgran$ & equigranular & A statement is equigranular to another $\stmt_1 \eqgran \stmt_2$ if they provide the same level of detail \\ 
		\hline 
		$\mu_{\stmt[u]}(\stmt)$ & Measure & Quantifies the precision of a statement $\stmt$ in terms of a reference unit statement $\stmt[u]$; it quantifies how many evolutions are compatible with $\stmt$ compared to the evolutions compatible with $\stmt[u]$  \\ 
		\hline 
		$S_{\stmt[u]}(\stmt)$ & Evolution entropy & The number of bits that separate the level of description of $\stmt$ from the one provided by the unit statement $\stmt[u]$; it quantifies the number of questions needed to go from the level of description of $\stmt$ to the level of description of $\stmt[u]$  \\ 
		\hline 
		$P(\stmt_2 | \stmt_1)$ & Probability & Quantifies quantifies the ratio of evolutions compatible with $\stmt_1$ that are also compatible with $\stmt_2$  \\ 
		\hline 
		 & Comparable & Two statements are comparable if one is finer than the other \\ 
		\hline 
		 & Finitely comparable & Two statements are finitely if the measure of one in terms of the other is finite and non-zero \\ 
		\hline 
	\end{tabular} 
	\caption{Granularity, measures and probability}
	\label{table:states}
\end{table}

\subsection{Systems, equilibria and state spaces}

So far we have made no guarantees as to whether each $X_t$ contains descriptions that refer to the same physical object. So we need to understand what is needed to be able to define a system and its state space. Intuitively, the state space should provide us with all the descriptions of the system at any point in time for any process. We are therefore asked to create an abstraction that stitches together all possible description, all possible snapshots, about the system and only about the system. There are therefore a few issues we need to consider.

First of all, when we say ``all process'' we can't really mean absolutely all of them, only the ones that somehow do not disrupt the system. For example, we can study the motion of a balloon on earth at 20 Celsius. This would not be possible on the surface of the sun as our balloon would not exist. The process has to be such that the system is recognizable to be the same, our descriptions must still make sense. We can imagine a space $X$ of all statement fragments like \statement{the position of the ball is between 2 and 3 meters} and a surjective map $\iota : X \times T \twoheadrightarrow X_t$ that adds \statement{at time t} for each possible time. The map $\iota$ must be a surjection that preserves the logical structure (i.e. logical operations, narrowness, compatibility, ...) so that all descriptions are only about the system and the system determines the basic logical relationships that are valid at all times (i.e. \statement{the position of the ball is between 2 and 3 meters} $\narrower$ \statement{the position of the ball is between 0 and 300 meters} at every $t$).

Note, though, that $\iota( \cdot, t_0)$ in general is not an isomorphism (i.e. a bijection that preserves the logical structure) for two reasons. The first is that not all states will be available at every moment in every process. As we said before, if the process is dissipative the set of possible states will become smaller. In that case, $\iota$ will map all the inaccessible states to $\contradiction$, as they will be impossible. Therefore $\iota$ is not in general a bijection. The second reason is that the statements in $X$ represent only descriptions about the system itself and nothing else. In a particular process, however, knowing the state of one system might tell us something about other system as well. The presence of correlations/coupling may mean that the precise knowledge of the degrees of freedom of the system may allow us to tell something about the environment or the internal structure (i.e. microstate). Mathematically, $\iota$ may map to statements that are narrower than the original (e.g. we can have $\iota($\statement{position of A is 1 meter}$, t_0) \equiv$ \statement{at time $t_0$ the position of A is 1 meter and the position of B is -1 meter}).

States coincide with snapshots of the system only in the case that the system is uncorrelated with environment and it does constrain the internal dynamics. This is also the case when the statements are broadest, which means it the case when the process entropy is maximized. We therefore define the \textbf{state entropy} as the maximum of all possible evolution entropies one can find for the given description of the system in all possible processes at all times.

In this context, entropy maximization is a necessary step to be able to define a system. If we are not able to find a process that, at least in some cases with good approximation, makes the evolution of the system independent from the environment and from the internal dynamics, we cannot talk about a system to begin with. In other words, if we are not able to isolate the system so that it can be studied independently, we are not going to be able to assign it a state, we are not able to construct a state space.

It is then useful to start studying pro

The statements present in our state space $X$, then, represent theIf $\iota$ can map to statements of different granularity, we cannot assign a unique evolution entropy to the statements of $X$. Yet,

 In our framework, we need some type of homomorphism between $\tdomain_t$ and $\tdomain_{t + \Delta t}$ that preserves narrowness and fineness, that tells us the system is the same.

So far we have only characterized how descriptions at different times can be related to each other. We now want to group descriptions that belong together, that describe a single system as it is subjected to different processes. In other words: we want to start talking about \textit{states of the system}, not just descriptions or snapshots. In the simplest form, we prepare the system in an initial state, let it evolve and look at the final state. We want to characterize the state space of all possible initial and final states under all process and characterize the process. Our framework already tells us that if the process is deterministic, the process can be characterized as a map $x(t+\Delta t) = f(x(t))$. If the process is non-deterministic, the evolutions for each state may split, and the probability $P(x(t + \Delta t) | x(t))$ can be used to characterize the fraction of evolutions that start in $x(t)$ and end in $x(t+\Delta t)$. Yet, it also tells us there are a few things that we have to worry about.

Second we want to make sure that the state space of the system really tells us only about the system itself. Within a specific process, for example, saying \statement{at time $t$, the velocity of object $A$ is $v$} may also mean that \statement{at time $t$, the velocity of object $B$ is $v$} because they are tied together. We say two systems are \textbf{independent} at a particular time if the description of one system does not constrain the description of the other. In terms of measure $\mu_{\stmt[u]_a \AND \stmt[u]_b}(x_a \AND x_b) = \mu_{\stmt[u]_a}(x_a) \mu_{\stmt[u]_b}(x_b)$ must factorize. If we want our state space to describe only the system at hand, then we are assuming that states are defined on the assumption that the system is independent at least from all other systems with the chosen processes.

On the other side, a state of a system is never really the description of the whole system, but of the part that is relevant to the process at hand. If we study the motion of the balloon, position and velocity may be enough; if we study it as a thermodynamic system, pressure and volume may be enough. The idea is that the state captures the system at a particular level of granularity and the process does not depend, is not sensitive, to the finer dynamics. We say a system is in \textbf{equilibrium} if the state tells us nothing about the

We stress that the independence cannot be, in general, independent throughout the whole process or there cannot be interactions between the systems. By the same token, the finer dynamics may play a part throughout the process. The constraint is only on the initial and final states, where the state space is being constructed. We also stress that these constraints make the state definition contingent upon the type of processes one studies. If we want to study a process where two systems are not independent, one that depends on the finer dynamics, our state definition will have to change to accommodate.

 In all processes we are considering, then, the system must be indAny other system would be independent If we found other systems that were always correlated For example, if  There can't be other systems that participate in the process whose state

We should caution that

The simplest setting is the follow

 into Our goal now is to construct state spaces for a system.

 Consequently any statement that corresponds to finitely many evolutions (e.g. \statement{the full trajectories is either $y=t \cdot 1 m/s$ or $y=-t \cdot 1 m/s$}) will be comparable to all other statements.

Also note that in our ideal gas example one of the quantities was determined with infinite precision. If we gave a range for all quantities, we would compare areas in that space, which we can always do. Statements with finite precision measurements are the only ones we can experimentally verify. So, we will require that all experimentally verifiable statements are comparable. We call \textbf{experimental domain} for the process the subset $\edomain[P] \subseteq \tdomain[P]$ of statements that can be, at least in line of principle, experimentally verified.\footnote{Yet, as we will see later, the precision of measurements will be quantified in terms of number of evolutions, not the range/uncertainty on the variables.}

While we leave the full mathematical characterization in the mathematical section, there are a few things of note. Verifiable statements will correspond, in terms of sets of $E$, to finite and infinite volumes. Requiring those to be comparable means that $E$ is a measurable space and in phase space, for example, the standard measure will be recovered. As the process domain corresponds to a $\sigma$-algebra on $E$, the experimental domain corresponds to a topology on $E$. Therefore notions of continuity can be recovered as well. The process domain is the closure of the experimental domain under standard Boolean operations. This means that all the process domain is fully specify by, at least in principle, experimentally verifiable statements. The space does not contain any assertion that cannot ultimately be defined in terms of experiments. Lastly, requiring that, given enough time, each statement will be verified means the experimental domain must be generated by countably many statements. This means the topology is second countable and one is limited to work with ``nice'' spaces (e.g. spaces of continuous functions, separable Hilbert spaces, ...) not because they are ``nice'', but because they are the only physically meaningful ones.

To sum up, we have two foundational object. The first object is the process domain $\tdomain[P]$ which is a Boolean algebra that contain all the possible description of our process. A subset $\edomain[P]$ contains those that are experimentally verifiable, which contain all the information in the full domain. The narrowest statements in $\tdomain[P]$ correspond to the evolutions $E$, that provide the full description at all times. The second object is the operator $\finer$ which gives us a way to compare two statements and decide which one gives a more refined description. Though two statements will not in general be comparable, all verifiable statements are comparable and all evolutions provide an equally refined description.

This framework imposes very few requirements which are of very general applicability. It does not matter what specific quantities are used, or whether the process is from physics, chemistry, control systems, ecology, economics, sociology, or some other realm we have yet to discover. In the end, if we have a process we need to model, our theory will need at least to provide those objects. This means that the results we find will have broad applicability as well.

\section{States}

As the process extends over time, it is natural to want to organize the statements in our process domain into partial domains, one for each moment in time. Given our time parameter $t \in T \subseteq \mathbb{R}$, we can imagine to carve a subdomain $\tdomain_t$ that would correspond to all the statements about the system one can make at the given time. Any logical operation on statements at fixed $t$ is also another statement at fixed $t$ therefore $\tdomain_t$ is a $\sigma$-complete Boolean sub-algebra of $\tdomain[P]$. The set $X_t \subset \tdomain_t$ of the narrowest statements will give the most precise description at that time: these are the possible states of our system.

We should note that, technically, the statements are not defined at an infinitesimal moment $t$, rather within a finite interval $[t, t + \Delta t]$. This represent the time-scale at which the process will be described, meaning the description within $\tdomain_t$ should not be sensitive to what happens at faster scales. The quantities used to represent states at time $t$, then, should really be thought as averages within $\Delta t$.

We should also note that there is nothing at this point requiring us that the set of descriptions $\tdomain_t$ and states $X_t$ will be the same at all times.

What we want to do now is relate states at different times. In particular we want to formally define the notion of deterministic process: one in which, given the state at a given time, we can always predict the state at all future times. This can be done through the use of the narrowness operator: $\stmt_{t_1} \narrower \stmt_{t_2}$ means that all evolutions compatible with $\stmt_{t_1}$ are also compatible with $\stmt_{t_2}$. In other words, if we find $\stmt_{t_1}$ to be true we will also find $\stmt_{t_2}$ to be true as well. Therefore a process is deterministic if, for all $t_1, t_2 \in T$, given $x_{t_1} \in X_{t_1}$ we can find $x_{t_2} \in X_{t_2}$ such that $x_{t_1} \narrower x_{t_2}$.

At each moment, we have a set of possible statements about the system corresponding to physically measurable properties. We call state the maximal description that can be given at a particular time. Given the state, we know all that can be said about the system in that moment. We define an evolution of the system as the maximal description of the system at all times. If we knew the exact evolution, we know all possible descriptions of the system at all times. Therefore picking an evolution is equivalent to picking a state at each time. A particular process will constrain how the system can evolve, it will constrain the possible evolutions. Therefore we formally define a process as a set of possible evolutions, the ones that are allowed by the process.

>>>>

The overall goal is to describe a generic system as it evolves in time by defining a minimal set of concepts that such a setup must define, and understand the tacit assumptions underlying state spaces and evolution laws. We want to find notions that can be applied in different settings, being it thermodyamics and/or dynamical systems alike. Then we want to derive general results based on such definitions. Here we provide an overview of the main features, leaving the formal definitions in the (yet to be written) appendix.

@@

Our starting point is the \textbf{process domain} $\tdomain[P]$, which is the set that contains all possible descriptions of the system at all levels of detail at all possible times. Depending on the process, it will contain statements like \statement{the average volume between 1 and 2 seconds is between 3 and 4 liters} or \statement{the trajectory of the particle is $y=10 \, m - t^2 \cdot 9.80665 \, m/s^2 $}. It does not matter at this point what the statements themselves are, except they must be, at least in line of principle, defined from experimentally well-defined starting points. Mathematically, $\tdomain[P]$ is a $\sigma$-complete Boolean algebra\footnote{A set of statements that is closed under negation (NOT), countable disjunction (OR) and countable conjunction (AND).} and it is the closure of the set of experimentally verifiable statements $\edomain[P] \subseteq \tdomain[P]$.\footnote{The set $\edomain[P]$ will form a Heyting algebra which, to be experimentally reachable, must allow a countable basis. The closure $\tdomain[P]$ is in terms of negation and countable disjunction.}

@@

Finally, we want to capture whether the truth of one or more statements incluences the truth of others. For example, we want to say that \statement{the temperature of the water in the glass is 3.98 C} is \textbf{independent} from \statement{the volume of the water in the glass is 10 $cL$} but is not independent from \statement{the density of the water in the glass is 1 $g/cm^3$}. We note $\stmt_1 \indep \stmt_2$ when two statements are independent.

@@

\begin{table}[h!]
	\centering
\begin{tabular}[h]{|c|c|p{6cm}|}
	\hline 
	Symbol & Name & Meaning \\ 
	\hline 
	$\tdomain[P]$ & Process domain & The set of all possible descriptions at all times and at all levels of granularity \\ 
	\hline 
	$E \subset \tdomain[P]$ & Possible evolutions & The set of descriptions that give a full account of the process; they correspond to the narrowest statements within $\tdomain[P]$ \\ 
	\hline 
	$A(\stmt)$ & Compatible evolutions & The set of evolutions that are compatible with the statement $\stmt$ \\ 
	\hline 
	$\narrower$ & narrower & A statement is narrower than another $\stmt_1 \narrower \stmt_2$ if the second one is true whenever the first one is \\ 
	\hline 
	$\comp$ & compatible & A statement is compatible with another $\stmt_1 \comp \stmt_2$ if they can be true at the same time \\ 
	\hline 
	$\indep$ & independent & Two statements are independent  $\stmt_1 \comp \stmt_2$ if the truth of one does not influence the truth of the other \\ 
\hline 
\end{tabular} 
	\caption{Process descriptions and their logical relationships}
	\label{table:logic}
\end{table}

@@

As the process extends over time, we want to organize the statements in our process domain into partial domains, one for each moment in time. Given a \textbf{time parameter} $t \in T \subseteq \mathbb{R}$, we can imagine to carve a \textbf{system domain} $\tdomain_t$ that would correspond to all the statements about the system one can make at the given time. In particular, we define a \textbf{one-step processes} one that only has an \textbf{initial domain} $\tdomain_{t_0}$ and a \textbf{final domain} $\tdomain_{t_1}$. Note that $\tdomain_t$ is a $\sigma$-complete Boolean sub-algebra of $\tdomain[P]$ since any logical operation on statements at fixed $t$ is also another statement at fixed $t$. The set $X_t \subset \tdomain_t$ of the narrowest statements will give the most precise description at that time, we call these the  possible \textbf{snapshots} of our system at that time.

@@

We should also note that the set of possible snapshots $X_t$ may not be the same at all times and may not correspond to the state of the system. First of all $X_t$ will only include those configurations that are compatible with at least by one possible evolution (i.e. $A(x) \neq \emptyset$ for all $x \in X_t$), which may change in time. For example, the set of possible configurations for a system under dissaptive processes will shrink as they converge to equilibrium. In quantum mechanics, the possible configurations after a measurement are restricted to the eigenestates, and a different choice of measurement will lead to a different process with different possible evolutions. Secondly, the best possible description may be broader than the full state of the system (e.g. when external interference forces us to give only a statistical account within our $\Delta t$) or narrower (e.g. when there are known correlations to the other degrees of freedom of the environment or of the system itself). In other words, the state space of the system (i.e. the set of all possible complete descriptions of the system and only about that system, independently of the process at hand) is something we'll construct later.

We say a process is \textbf{deterministic} over the system domains $\{X_t\}_{t \in T}$ if given the snapshot at one time we can always predict the snapshot at all future times. Formally, let $t_1, t_2 \in T$ such that $t_1 \leq t_2$, for any $x_1 \in X_{t_1}$ we can find an $x_2 \in X_{t_2}$ such that $x_1 \narrower x_2$.

@@

Conversely, we say a process is \textbf{reversible} over the system domains $\{X_t\}_{t \in T}$ if given the snapshot at one time we can always reconstruct the snapshot at all past times. Formally, let $t_1, t_2 \in T$ such that $t_1 \leq t_2$, for any $x_2 \in X_{t_2}$ we can find a $x_1 \in X_{t_1}$ such that $x_2 \narrower x_1$.

@@

If we limit ourselves to deterministic processes, with these few definitions, we have already found that the set of possible evolutions ``grows bigger'' under irreversible evolution while it remains the ``the same size'' under reversible evolution. This is the core idea of the second law.

@@

\begin{table}[h!]
	\centering
	\begin{tabular}[h]{|c|c|p{6cm}|}
		\hline 
		Symbol & Name & Meaning \\ 
		\hline 
		$t \in T$ & Time parameter & Time is treated as a real valued parameter within the range $T \subseteq \mathbb{R}$ \\ 
		\hline 
		$\tdomain_t \subset \tdomain[P]$ & System domain & The set of descriptions at a given time \\ 
		\hline 
		$X_t \subset \tdomain_t$ & Possible snapshots & The set of descriptions that give a full account at time $t$; they correspond to the narrowest statements within $\tdomain_t$ \\ 
		\hline 
		$x(t) \narrower x(t + \Delta t)$ & Determinism & Given the snapshot at one time we can predict the snapshot at future times \\ 
		\hline 
		$x(t) \narrower x(t - \Delta t)$ & Reversibility & Given the snapshot at one time we can reconstruct the snapshot at past times \\ 
		\hline 
	\end{tabular} 
	\caption{System domain, snapshots, determinism and reversibility}
	\label{table:states}
\end{table}

@@

Quantifying evolutions allows us to define probability as well. If we consider all the evolutions compatible with \statement{the coin is fair}, we must have that half of them are compatible with \statement{the coin is head} while the other are compatible with \statement{the coin is tails}.  Therefore we have that the \textbf{probability} $P(\stmt_2 | \stmt_1) = \mu_{\stmt_1}(\stmt_1 \AND \stmt_2)$ where $\stmt_1 \AND \stmt_2$ is the logical conjunction (AND) between the two statements. In other words, the probability quantifies the ratio of evolutions compatible with $\stmt_1$ that are also compatible with $\stmt_2$.

We can study a one-step process by tracking the probability $P(\stmt_1 | \stmt_0)$ that the final description $\stmt_1$ is true given that the initial description $\stmt_0$ is true, which is simply the fractions of evolutions starting in $\stmt_0$ (i.e. are compatible to $\stmt_0$)) that end in $\stmt_1$. The fact that \textbf{probability theory can be used to fully characterize all one-step processes}, the simplest processes, tells us why it is so important in physics.

@@

\begin{table}[h!]
	\centering
	\begin{tabular}[h]{|c|c|p{6cm}|}
		\hline 
		Symbol & Name & Meaning \\ 
		\hline 
		$\finer$ & finer & A statement is finer than another $\stmt_1 \finer \stmt_2$ if it provides a more refined description than the second \\ 
		\hline 
		$\eqgran$ & equigranular & A statement is equigranular to another $\stmt_1 \eqgran \stmt_2$ if they provide the same level of detail \\ 
		\hline 
		$\mu_{\stmt[u]}(\stmt)$ & Measure & Quantifies the precision of a statement $\stmt$ in terms of a reference unit statement $\stmt[u]$; it quantifies how many evolutions are compatible with $\stmt$ compared to the evolutions compatible with $\stmt[u]$  \\ 
		\hline 
		$S_{\stmt[u]}(\stmt)$ & Evolution entropy & Defined as $\log \mu_{\stmt[u]}(\stmt)$ represents the number of bits that separate the level of description of $\stmt$ from the one provided by the unit statement $\stmt[u]$; it quantifies the number of questions needed to go from the level of description of $\stmt$ to the level of description of $\stmt[u]$  \\ 
		\hline 
		$P(\stmt_2 | \stmt_1)$ & Probability & Defined as $\mu_{\stmt_1}(\stmt_1 \AND \stmt_2)$ quantifies the ratio of evolutions compatible with $\stmt_1$ that are also compatible with $\stmt_2$  \\ 
		\hline 
		 & Comparable & Two statements are comparable if one is finer than the other \\ 
		\hline 
		 & Finitely comparable & Two statements are finitely if the measure of one in terms of the other is finite and non-zero \\ 
		\hline 
	\end{tabular} 
	\caption{Granularity, measures and probability}
	\label{table:states}
\end{table}

\subsection{States and system independence}

So far we have made no guarantees as to whether each $\tdomain_t$ contains descriptions that refer only to a particular physical object. So we need to understand what is needed to be able to define a system and its state space. Intuitively, the state space should provide us with all the descriptions of the system, and only of the system, at any point in time for any process. We are therefore asked to create an abstraction that stitches together all possible descriptions, all possible snapshots, about the system and only about the system, which means there are a few issues we need to consider.

First of all, when we say ``all processes'' we can't really mean absolutely all of them, but only the one for which the system is well defined. For example, we can study the motion of a balloon on earth at 20 Celsius. This would not be possible on the surface of the sun as our balloon would not exist. The process has to be such that the system is recognizable to be the same, our descriptions must still make sense. We can imagine a \textbf{state domain} $\tdomain$ of all statement fragments like \statement{the position of the ball is between 2 and 3 meters} together with a surjective map $\iota : \tdomain \times T \twoheadrightarrow \tdomain_t$ that adds \statement{at time t} for each possible time. The map $\iota$ must be a surjection that preserves the logical structure (i.e. logical operations, verifiability, narrowness, compatibility, ...) so that the basic logical relationships determined by the system itself are are valid at all times (i.e. \statement{at time t the position of the ball is between 2 and 3 meters} $\narrower$ \statement{at time t the position of the ball is between 0 and 300 meters} for every $t$).\footnote{We stress that the mathematical requirements are necessary but not sufficient. While the same system must map to the same logical structure but the same logical structure does not necessarily map to the same system (e.g. two distinct system can be represented by the same mathematical structure).}

Note, though, that $\iota( \cdot, t_0)$ in general is not an isomorphism (i.e. a bijection that preserves the logical structure) for two reasons. The first is that not all states will be available at every moment in every process. As we said before, if the process is dissipative the set of possible states will become smaller. In that case, $\iota$ will map all the inaccessible states to $\contradiction$, as they will be impossible. Therefore $\iota$ is not in general a bijection. The second reason is that the statement fragments in $\tdomain$ represent only descriptions about the system itself and nothing else. In a particular process, however, knowing the state of one system might tell us something about other system as well. The presence of correlations/coupling may mean that the precise knowledge of the degrees of freedom of the system may allow us to tell something about the environment or the internal structure (i.e. microstate). Mathematically, $\iota$ may map to statements that are narrower than the original fragment (e.g. we can have $\iota($\statement{position of A is 1 meter}$, t_0) \equiv$ \statement{at time $t_0$ the position of A is 1 meter and the position of B is -1 meter}).

In physics terms, the construction of the state domain requires us to be able to identify and separate the system from all other systems and the environment. We can talk about a book and its state because we can, in many situations, describe and manipulate the book without describing or manipulating other books, the bookcase or the table. If $\tdomain_{t}$ is the domain of our system at time $t$ and $\tdomain[O]_t$ is the domain for another system, we say that the two domains are \textbf{independent} if, taken any two statements $\stmt_d \in \tdomain_{t}$ and $\stmt_o \in \tdomain[O]_{t}$ are independent. In this case there are no correlations: every possible pair of snapshots of each domain is a possible snapshot for the combined system. Therefore if we take $\stmt[u]_d \in \tdomain_{t}$ and $\stmt[u]_o \in \tdomain[O]_{t}$ to be unit statements for the respective measure, we have:
\begin{equation}\label{entropy_sums}
\begin{aligned}
	\mu_{\stmt[u]_d \AND \stmt[u]_o}(\stmt_d \AND \stmt_o)=\mu_{\stmt[u]_d}(\stmt_d) \mu_{\stmt[u]_o}(\stmt_o) \\
S_{\stmt[u]_d \AND \stmt[u]_o}(\stmt_d \AND \stmt_o)=S_{\stmt[u]_d}(\stmt_d) + S_{\stmt[u]_o}(\stmt_o)
\end{aligned}
\end{equation} 
That is, \textbf{for independent systems the process entropy can be seen as the sum of the entropy of each system}.\footnote{Note the direction in which the construction unfolds. We are not defining the entropy on the single systems and then, if they are isolated, the entropy sums. It's the opposite: the entropy is initially defined on the joint system for a particular process as the evolutions describe all systems. If the system is independent of all others, the evolutions are grouped such that their measure factorizes, and we can assign a unique measure to each statement of the independent system, as this will not depend on the description of other systems. If we cannot find this situation, the entropy of the system cannot be defined uniquely on the separate system, but only on the joint. In physics terms, we are unable to treat the description as set of independent degrees of freedom (e.g. describing only the pressure of a gas without the volume, or only the position without the momentum).}

We say we have an \textbf{independent system} if the system is independent from all other systems. In this case, the snapshot is a description of the system and only of the system. The state domain, then, represent descriptions of the system when it is independent. A \textbf{state} $x$ of a system is a snapshot in these conditions and the \textbf{state space} is the set of all possible states.\footnote{There is still the added complication that, even if the system is independent, not all states may be accessible. For example, the thermodynamic states may change depending on the constraint applied to the system (e.g. presence or absence of a magnetic field); in quantum mechanics, the accessible states after a spin measurement will only be the ones along the measured direction. So, in general, the state space stitches together all the possible independent snapshots at all times over all possible processes.} In the same way, the process entropy becomes a property of the system itself and we can define the \textbf{state entropy} $S_{\stmt[u]} : X \to \mathbb{R}$ as the process entropy in these conditions, where $\stmt[u] \in X$ is a state taken as reference. That is:
\begin{equation}
	S_{\stmt[u]}(x) = S_{\iota(\stmt[u], t_0)}(\iota( x, t_0))_{indep}
\end{equation}

We say $\xi^a : X \to \mathbb{R}$ are a set of state variables if they uniquely identify a state. That is, we can find an isomorphism (a bijection that preserve the logical structure) $\phi : \mathbb{R}^n \to X$ such that $\phi(\xi^a(x)) = x$ for all $x \in X$. Then each state space comes equipped with a \textbf{state function} $S=S(\xi^a)$. We'll see that, for thermodynamic systems, this will correspond to the usual thermodynamic entropy. We define a \textbf{purely mechanical} system one for which the state entropy is the same across all states (i.e. $S=k$ where $k$ is a constant, typically chosen to be zero). We'll see that classical phase space and quantum state spaces are state spaces of purely mechanical systems.

Since we can only properly talk about a system when it is independent from the other, what can we do in the other cases? The idea that we can still talk about a system in that case if we are able, at least in line of principle, to decouple the system from all others while not changing the nature of the system itself. For example, we can talk about a molecule of a gas because we have a way to take such molecule out of the gas. The construction of a state space of a system, then, also requires telling us what all the possible correlated description will go when the system is made independent. Mathematically, we need a deterministic one-step process that takes any initial snapshot of the system to a final state, and does so without disrupting the system (i.e. the relaxation process must take each state to itself). We call this process \textbf{relaxation}.\footnote{The state space is invariant over the relaxation. In other word, the state space is a symmetry of the relaxation process.}

If $x_0$ is the initial snapshot and $x_1$ is the final state, given that the relaxation process is deterministic, we have $x_1= x_1(x_0)$ and 
\begin{equation}
S_{\iota(\stmt[u], t_1)}(x_0) \leq S_{\iota(\stmt[u], t_1)}(x_1) = S_{\iota(\stmt[u], t_1)}(\iota( x, t_1)) = S_{\stmt[u]}(x)
\end{equation}
The process entropy increases and is maximized when the system is independent.\footnote{Mathematically, the relaxation process in needed to compare the process entropies for the correlated descriptions. As the same description can be more or less correlated in different cases, the independent ones work as a reference and we need a way to relate to them.}

Note that all snapshots that relax to the same state will have a process entropy lower or equal to the state entropy. Therefore \textbf{the state entropy is the largest possible process entropy} for which the description given by the state applies. This does not simply represent a subjective information theoretic consideration (e.g. we know only about the system); it is a physical consideration (e.g. knowledge about one system tells us nothing about the others). As we make the system more independent, there are more possible evolutions the combined system can have. Therefore in this framework \textbf{entropy maximization is equivalent to requiring system independence}. Entropy maximization, more than a law of physics, is a requirement to properly define the state space of a system, a requirement for doing physics. As such, we are therefore biased towards looking for such situations.

\subsection{Thermodynamic processes}

% DO WE NEED INTERNAL STATE? On the other side, a state of a system is never really the description of the whole system, but of the part that is relevant to the process at hand. If we study the motion of the balloon, position and velocity may be enough; if we study it as a thermodynamic system, pressure and volume may be enough. The idea is that the state captures the system at a particular level of granularity and the process does not depend, is not sensitive, to the finer dynamics. We say a system is in \textbf{equilibrium} if the state tells us nothing about the
